Research Experience

### 2010-06-23 doc/thesis/thesis.tex
sort of done with example circuit particle filter.

\section{Unscented Kalman Filter}
The unscented kalman filter is a powerful bayes-gaussian filter that attempts
to model the posterior distribution of dynamical systems as a multivariate
Gaussian. The Unscented Kalman Filter (UKF) generalizes the Extended Kalman
Filter by allowing the state update to be a function, $g$,

@@

In order to estimate the posterior at $t$, a deterministic set of points (often 2 
per dimension, plus 1 at the mean)
weighted according to a Gaussian estimate of $X(t-1)$ are passed through
the update equation. This set of points are then used to estimate the 
mean and covariance of $X(t)$. The benefit of this, is that it requires
no Jacobian and only a few extra calculations to get a decent esimate of
a posterior Gaussian. In the BOLD case, the set of equations we are modeling have no 
closed form solution, and finding the Jacobian is impossible without approximations. Although 
\cite{Riera2004} claims to use a Jacobian for their solution, they are
in fact using the Jacobian of $\dot{s}$, $\dot{q}$, $\dot{f}$, $\dot{v}$,
rather than $s$, $q$, $f$, and $v$. This is important because the Extended
Kalman filter depends on the jacobian to map a gaussian through the 
advancement of time. Thus the popular Extended Kalman Filter won't work
in this case, whereas the Unscented Kalman Filter still does.

The difficulty with the Kalman filter is that it is not well suited
to parameter estimation. If we wanted to find the mean of a static
parameters, such as $V_0$, it would be necessary to guess a mean 
and variance for all time. This is because the Kalman filter is primarily
concenrned with estimating the output, and is not really concerned
with estimating the underlying state, $X(t)$. As such estimated 
output, $\hat{Y}(t)$, is simply a weighted sum of the measurement,
$Y(t)$, and the expected value of the output based on the propagated
state variables, $\hat{X}(t)$. As such, when the  the filter will attenuate measurements
the vastly disagree with $E[h(\hat{X}(t))]$, but eventually

The difficulty
of using a Kalman Filter, however, is that it assumes a multivariate 
Gaussian for the state variables, $X(t-1)$. The problem with this is that 
when a system is significantly nonlinear, the state at 
$X(t)$ will almost certainly be non-Gaussian, and thus estimating
$X(t+1)$ with $X(t)$ as a multivariate Gaussian will perpetually introduce
error in the distribution. Furthermore, it is not really known what 
sort of underlying distributions may exist in such a mixed biological,
mechanical, chemical system such as the brain. Assuming the parameters
all to be Gaussian may in fact be a gross error. On the other hand, if
the our tests indicate a Gaussian distribution is a good fit, it would
be a wise plan to go back and consider the use of a modified Unscented 
Kalman Filter rather
than the full Particle Filter we discuss in the next section.

\section{Proposed Approach}
\subsection{Goal}
The ultimate goal of this project is to provide a new set of tools
for analyzing FMRI data. Whereas SPM techniques have been highly 
successful at finding macroscopic regions of activation, linear 
modeling can carry significant bias error due to lack of model
flexibility. While adding parameters can significantly increase
error due to model variance, this effect is mitigated by the fact
that we plan to use a model that is based on first principals. The
purpose of this paper is thus to evaluate the potential of using
a particle filter along with the BOLD model to derive physical 
parameters. In so doing, we hope to be able to show that neuronal
efficacy, $\epsilon$ is a suitable variable for estimating voxel 
activation from a standard FMRI image. We also hope to show that 
estimated posterior distribution of the parameters, derived from
the particle filter, is able to provide an accurate measure of the
confidence interval.

@@

Although not generally necessary for particle filters, we will make a few
assumptions based on the particular type of systems faced in biological 
processes. First, the systems are assumed to be time invariant. This 
assumption is based on the idea that if you froze the system for $\Delta t$
seconds, when unfrozen the system would continue as if nothing happend. 
Few biological systems are predictible enough for them to be summarized
by a time varying function. Although the heart may seem like an obvious
exception, period between heartbeats vary often enough that prediction
would necessate another state-space model. In short, we
assume no parameters are time varying, because not enough information exists to
describe any of theme in that way. Luckily particle filters are capable 
of dealing with non-white, non-Gaussian noise, so unanticipated influence
may be re-factored as noise. Secondly we assume that input cannot directly
influence the output, which in the case of the BOLD signal is a good assumption.
%Third, through some sort of preprocessing, we will assume that $\nu_d$ can be
%decreased or completely removed, since it 
Third, we will assume noise is additive, and that $\nu_x$ may be projected into
a weiner, or other summing process that is additive with $g$ and $\nu_y$, which
will be named $\nu_d$.
Finally, $x(t)$ will encapsulate $\theta$, the unknown model constants, which
means that the vector $\dot{x}$ will always have members
that are 0. The results of these assumptions are a simplified version of the
state space equations:

@@

\section{Simple Example}
So a simple example may be used to demonstrate the proposed use of a particle filter.
Thus, let us consider a simple half wave rectifier circuit, shown in figure 
\ref{fig:HalfWaveRectifier}

\begin{figure}
\begin{center}
\begin{circuitikz}[scale=2, american]
\draw
 (0,0)  node[transformer core] (T) {}
 (T.A1) -- (-1,0)
 (T.A2) -- (-1,-1.05)  to[V, v=$u(t)$] (-1, 0)
 (T.B1) -- (.5, 0) to[D, l=$v_t$] (1.5,0) to[C=$C$] (1.5, -1.05)
 (1.5, 0) -- (2.5, 0) to[R=$Rm$, v=$v_y$] (2.5, -1.05) -- (T.B2) 
 (T.base) node {G};
\end{circuitikz}
\label{fig:HalfWaveRectifier}
\caption{An Example Half Wave Rectifier Circuit, where $G$ is the transformer
gain, $v_t$ is the activation voltage of the diode, $u(t)$ is the input at time $t$, 
$C$ is the capacitance, $R$ is the load resistance and $v_y$ is the output voltage}
\end{center}
\end{figure}

Thus when $u(t)G$ is less than $v_t$, the circuit will discharge the capacitor,
but when $u(t)G$ is greater than $v_t$, the capacitor will charge up. Unlike
the normal rectifier circuit, we'll deal with a generic input
Let $V_C$ be the

>>>>

\section{Unscented Kalman Filter}
The unscented kalman filter is a powerful gaussian/bayes filter that attempts
to model the posterior distribution of dynamical systems as a multivariate
Gaussian. The Unscented Kalman Filter (UKF) generalizes the Extended Kalman
Filter by allowing the state update to be a function, $g$,

@@

In order to estimate the posterior at $t$, a deterministic set of "sigma" points 
(often 2 per dimension, plus 1 at the mode of the multivariate distribution)
weighted according to a Gaussian estimate of $X(t-1)$ are passed through
the update equation. This set of points are then used to estimate the 
mean and covariance of $X(t)$. The benefit of this, is that it requires
no Jacobian and only a few extra calculations to get a decent esimate of
a posterior Gaussian. In the BOLD case, the set of equations we are modeling have no 
closed form solution, and finding the Jacobian is impossible without approximations. Although 
\cite{Riera2004} claims to use a Jacobian for their solution, they are
in fact using the Jacobian of $\dot{s}$, $\dot{q}$, $\dot{f}$, $\dot{v}$,
rather than $s$, $q$, $f$, and $v$. This is important because the Extended
Kalman filter depends on the jacobian to map a gaussian through the 
advancement of time. Thus the popular Extended Kalman Filter won't work
in this case, whereas the Unscented Kalman Filter still does.

\begin{figure}
\label{fig:EKFWorking}
\caption{Examples of update distributions, using an Kalman Filter, \cite{Thrun}}
\end{figure}

The difficulty
of using a Kalman Filter, however, is that it assumes a multivariate 
Gaussian for the state variables, $X(t-1)$. The problem with this is that 
when a system is significantly nonlinear, the state at 
$X(t)$ will almost certainly be non-Gaussian, and thus estimating
$X(t+1)$ with $X(t)$ as a multivariate Gaussian will perpetually introduce
error in the distribution. Furthermore, it is not really known what 
sort of underlying distributions may exist in such a mixed biological,
mechanical, chemical system such as the brain. Assuming the parameters
all to be Gaussian may in fact be a gross error. On the other hand, if
the our tests indicate a Gaussian distribution is a good fit, it would
be a wise plan to go back and consider the use of a modified Unscented 
Kalman Filter rather
than the full Particle Filter we discuss in the next section. These are 
non-trivial issues given that the assumption of Gaussianity is what allows
the UKF to estimate the posterior using only the first and second moments;
two parameters that don't uniquely describe most distributions.

Another potential problem with the UKF is that typically the sigma points,
used to estimate the posterior probability, $P(X(t) | X(t-1), u(t))$, are
located on the main axes. As a result, the covariances are not allowed to 
be effected by the state transition the same way the variances are. While
this may be reasonable in low-dimensional systems, high demsional systems
have a much greater potential for interplay between the variables. While
this problem is relatively easy to fix (by using more sigma points off
the main axes), there is a definite cost in complexity.

Ultimately, the nonlinearities of the BOLD model make Gaussian estimates
too unrealistic for practical use. As \ref{fig:Transformation} shows, even 
holding all other parameters steady, the nonlinearities due to $q$ and $v$
alone are enough to give a non-gaussian result from a two variables 
gaussian input.

\begin{figure}

\label{fig:Transformation}
\caption{a) A Two-Variables gaussian input of $q$ and $v$, b) The 
resulting distribution after being evolved for two seconds. c) The
estimated Gaussian based on the mean and covariance of b). d) The
residual between the two distributions.}
\end{figure}

\section{Proposed Approach}
\subsection{Goal}
The ultimate goal of this project is to provide a new set of tools
for analyzing FMRI data. Whereas SPM techniques have been highly 
successful at finding macroscopic regions of activation, linear 
modeling can carry significant bias error due to lack of model
flexibility. While adding parameters can significantly increase
error due to model variance, this effect is mitigated by the fact
that we plan to use a model that is based on first principals. The
purpose of this paper is thus to evaluate the potential of using
a particle filter along with the BOLD model to derive physical 
parameters. In so doing, we hope to be able to show that one or more
parameters are a suitable replacement for estimating voxel 
activation from a standard FMRI image. We also hope to show that 
estimated posterior distribution of the parameters, derived from
the particle filter, is able to provide an accurate measure of the
confidence interval.

@@

Although not strictly necessary for particle filters, we will make a few
assumptions based on the particular type of systems faced in biological 
processes. First, the systems are assumed to be time invariant. This 
assumption is based on the idea that if you froze the system for $\Delta t$
seconds, when unfrozen the system would continue as if nothing happend. 
Few biological systems are predictable enough for them to be summarized
by a time varying function. Even systems that one might assume work in that
way are actually much more complicated. The heart may seem like an obvious
exception, however, the period between heartbeats vary often enough that prediction
would necessitate another state-space model. In short, we
assume no parameters are time varying, because not enough information exists to
describe any of theme in that way. Luckily particle filters are capable 
of dealing with non-white, non-Gaussian noise, so unanticipated influence
may be re-factored as noise. Secondly we assume that input cannot directly
influence the output, which in the case of the BOLD signal is a good assumption.
%Third, through some sort of preprocessing, we will assume that $\nu_d$ can be
%decreased or completely removed, since it 
Third, we will assume noise is additive, and that $\nu_x$ may be projected into
a Weiner, or other summing process that is additive with $g$ and $\nu_y$, which
will be named $\nu_d$.
Finally, $x(t)$ will encapsulate $\theta$, the unknown model constants, which
means that the vector $\dot{x}$ will always have members
that are 0. The results of these assumptions are a simplified version of the
state space equations:

@@

\section{Simple, Nonlinear Example}
A simple example may be used to demonstrate the proposed use of a particle filter.
Thus, let us consider a simple half wave rectifier circuit, shown in figure 
\ref{fig:HalfWaveRectifier}

\begin{figure}
\begin{center}
\begin{circuitikz}[scale=2, american]
\draw
 (0,0)  node[transformer core] (T) {}
 (T.A1) -- (-1,0)
 (T.A2) -- (-1,-1.05)  to[V, v=$u(t)$] (-1, 0)
 (T.B1) -- (.5, 0) to[D, l=$v_t$] (1.5,0) to[C=$C$] (1.5, -1.05)
 (1.5, 0) -- (2.5, 0) to[R=$Rm$, v=$v_y$] (2.5, -1.05) -- (T.B2) 
 (T.base) node {G}
 (T.B1) to[open, *-*, v=$V_1$] (T.B2); 
\end{circuitikz}
\label{fig:HalfWaveRectifier}
\caption{An Example Half Wave Rectifier Circuit, where $G$ is the transformer
gain, $v_t$ is the activation voltage of the diode, $u(t)$ is the input at time $t$, 
$C$ is the capacitance, $R$ is the load resistance and $v_y$ is the output voltage}
\end{center}
\end{figure}

The half wave rectifier circuit smoothes the gaps between high voltage
with a capacitor. Thu, when $u(t)G$ is less than $v_t$, the circuit will 
discharge the capacitor and maintain a non-zero voltage,
but when $u(t)G$ is greater than $v_t$, the output voltage will be set
by $u(t)G$ and the capacitor will charge up. We will assume a very simple
model for all the components, ignoring the complex nonlinear behavior
that can occur in a diode. Although rectifiers are typically thought
of as receiving an AC circuit 60 Hz, we will ignore such specifics and 
assume the voltage across the output of the transformer is simply a scalar
multiple of the input voltage.  As discussed in the \ref{Sec:ParticleFilter} any
variable with uncertainty must be part of the state variable. Therefore
the state variable will be: $X(t) = \{G, v_t, C, R_m, v_y\}$. Of course,
$u(t)$ cannot be allowed to be a square wave in such a system, since that
such a signal would never get across the transformer and regardless it
would necessitate an unrealistic infinite current across the capacitor.
The state equations would then be

\begin{equation}
v_y(t)  = f(v_y(t-1, u(t)) =  \begin{cases} 
        u(t)G & \text{ if }  u(t)G-v_y \ge v_t\\
        v_y(t-1)\left(1 - \frac{1}{R_mC}\right) & \text{ if }  u(t)G-v_y < v_t
    \end{cases} 
\end{equation}

To run the particle filter is relatively easy then, since there exists 
a recursive definition of the dynamic state variable, $v_y$. To start
with, an initial distribution must be assumed and while at first a 
Gaussian seems like a good idea, all the static state variables are strictly
positive and thus not well suited to the Gaussian. In this case then,
it would be wise to start with a Gamma distribution, and just be wary of
any standard deviation that gets larger than the prior mean. We will define
the gamma distribution as follows:

\begin{equation}
X \sim Gamma(k, \theta) \rightarrow f(x) = x^{k-1}\frac{e^{-x/\theta}}{\theta^k\Gamma(k)}
\end{equation}

where $\Gamma$ is the gamma function.
The in some ways the margin for error is decided by the weighting function, which
here will define as $W(V_y, v_y)$, where $V_y$ is the actual measurement, and 
$v_y$ is the estimate by a particular particle. The choice of this function is difficult,
and although the Gaussian is typically used, we found the exponential helpful
in dealing with particle deprivation.  The algorithm will then look like the following,

\begin{algorithmic}
\STATE Initialize $N_p$ Particles
\FOR{$i$ in $N_p$}
    \STATE $G \sim Gamma(\frac{\mu^2_G}{\sigma^2_G}, \frac{\sigma^2_G}{\mu_G})$
    \STATE $v_t \sim Gamma(\frac{\mu^2_{v_t}}{\sigma^2_{v_t}}, \frac{\sigma^2_{v_t}}{\mu_{v_t}})$
    \STATE $C \sim Gamma(\frac{\mu^2_C}{\sigma^2_C}, \frac{\sigma^2_C}{\mu_C})$
    \STATE $R_m \sim Gamma(\frac{\mu^2_R}{\sigma^2_R}, \frac{\sigma^2_R}{\mu_R})$
    \STATE $v_y = 0$, (Assume the system has been off for a long time)
    \STATE let $X_i(0) = \{G, v_t, C, R_m, v_y\}$
    \STATE let $w_i(0) = 1$ or to make a flat prior, $w_i(0) = \frac{1}{Pr(X_i(0))}$ 
\ENDFOR
\FOR{$t$ in Set of Measurement Times}
    \FOR{$i$ in $N_p$}
        \STATE $v_y(t) = f(v_y(t-1, u(t))$
        \STATE $w_i(t) = w_i(t-1)W(V_y(t), v_y(t))$
    \ENDFOR
\ENDFOR
